\newpage
{\Huge \bf Acknowledgements}
\vspace{24pt} 

I could not have done this dissertation without the support of my two wonderful advisors Pietro Lio and Jim Haseloff.
Pietro has been consistent encouragement throughout this entire process and he never ceases to amaze me with the depth of his knowledge and perpetual enthusiam.
Jim was extraordinarily kind to let a computer scientist run around his lab for a few months and I appreciate that he let me get my hands dirty.
Both of their groups have also been extremely welcoming despite me not being in either lab full time.

Emma Talbot was my mentor in the Haseloff group and taught me everything I needed to know about the wet lab in order to get this dissertation done on time.
In Pietro's lab, Marco Barsacchi provided extremely helpful advice on how to combine VAEs and FBA and was always willing to chat.
I really appreciate both of their help.

To my Cambridge friends Tom, Harriet, Emily, and Ian: thank you for putting up with my craziness and joining me for many meals during this entire process.
Seren, thank you for the copious amounts of pesto you made me as well as your emotional support throughout this process.

Thank you also to Juanky Perdomo and my dad, James Larus, for reading drafts of this dissertation and giving me feedback.

Finally, I would like to thank all of my friends and family who encouraged me on this journey and helped me get to where I am today.
In particular, I need to thank my parents, James Larus and Diana Stone, who have nurtured my love of learning and research and put me on the track that led me here. 

\newpage
\vspace*{\fill}